---
title: "The calibrate / simulate workflow in covsim"
author: "Njål Foldnes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The calibrate / simulate workflow in covsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Motivation

Since covsim 1.2.0, each of the three algorithms (`IG`, `PLSIM`, `VITA`) is
split into two steps:

1. **Calibration** — fits the model parameters that achieve the target
   covariance (and the target marginal moments where applicable). This step
   is the expensive one. It only needs to be run once for a given target.
2. **Simulation** — draws data from the calibrated model. Fast. Can be
   repeated as often as needed, returning a list of independent samples.

The original `rIG()`, `rPLSIM()` and `vita()` functions are unchanged and
still work; they are now thin wrappers that calibrate and simulate in one
call. New code that needs many simulated datasets from the same target
should call the calibrator once and reuse the calibration object.

## The general pattern

```{r eval=FALSE}
library(covsim)

# 1. calibrate once
cal <- calibrate_ig(sigma.target, skewness, excesskurtosis)
#   ... or calibrate_plsim(...)
#   ... or calibrate_vita(margins, sigma.target)

# 2. simulate one or more datasets
draws <- simulate(cal, nsim = 100, N = 1000)   # list of 100 data matrices
```

`simulate()` here is the standard `stats::simulate` generic; covsim
registers S3 methods for it on the three calibration object classes
(`covsim_calib_ig`, `covsim_calib_plsim`, `covsim_calib_vita`).

## IG (Independent Generator)

```{r}
library(covsim)
set.seed(1)
d <- 4
sigma.target <- cov(MASS::mvrnorm(50, mu = rep(0, d), Sigma = diag(d)))

cal_ig <- calibrate_ig(sigma.target,
                       skewness        = rep(1, d),
                       excesskurtosis  = rep(3, d))
class(cal_ig)

# draw five independent datasets from the same calibration
draws <- simulate(cal_ig, nsim = 5, N = 10000)
length(draws)
dim(draws[[1]])
round(cov(draws[[1]]) - sigma.target, 3)
```

The first call to `calibrate_ig` does the work; subsequent `simulate()`
calls are nearly free.

## PLSIM (Piecewise Linear)

PLSIM's pairwise calibration is the slow step (one nonlinear search per
variable pair). Splitting calibrate / simulate is especially useful here
because a simulation study often needs many replications under the same
target.

```{r}
set.seed(1)
sigma.target <- cov(MASS::mvrnorm(50, mu = rep(0, 4), Sigma = diag(4)))
cal_pl <- calibrate_plsim(sigma.target,
                          skewness       = rep(1, 4),
                          excesskurtosis = rep(3, 4),
                          verbose        = FALSE)

# Re-use the calibration for many replications
replications <- simulate(cal_pl, nsim = 50, N = 500)
length(replications)
```

## VITA (Vine-To-Anything)

`calibrate_vita()` calibrates a regular vine to match the target
covariance under user-specified marginals. The resulting calibration
object embeds the underlying `vine_dist`, so it can be passed to
`simulate()` (recommended) or to `rvinecopulib::rvine()` directly.

```{r eval=FALSE}
set.seed(1)
d <- 5
sigma.target <- cov(MASS::mvrnorm(50, mu = rep(0, d), Sigma = diag(d)))
margins <- lapply(sqrt(diag(sigma.target)),
                  function(s) list(distr = "norm", sd = s))

cal_vita <- calibrate_vita(margins, sigma.target,
                           Nmax    = 1e5,
                           cores   = parallel::detectCores(),
                           verbose = FALSE)

draws <- simulate(cal_vita, nsim = 10, N = 1000)
```

### Speed and parallelism

Since covsim 1.2.0, `vita()` calibration uses cached uniform inputs (via
`rvinecopulib::inverse_rosenblatt()`) so the inner root-finding objective
is deterministic; the previous multi-stage stochastic root search has
been replaced by a single `uniroot()` call. In addition, pair-copulas at
the same tree level are calibrated in parallel via `parallel::mclapply`.
Pass `cores = N` to use N workers (defaults to all detected cores).

The arguments `numrootpoints`, `conflevel`, and `numpoints` to `vita()`
are retained in the signature for backward compatibility but no longer
have any effect.

## When to use the wrappers vs the new API

Use the new API when:

- You need many replications from the same target (simulation studies).
- You want to inspect the calibrated model before simulating.
- You want to call `simulate()` in pipelines (it is the standard R generic).

Use the original `rIG()` / `rPLSIM()` / `vita()` wrappers when:

- You only need a one-shot draw.
- You're maintaining old code; both still work and produce the same
  output shape as in covsim 1.1.0.
